A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-05-05 paper/paper.tex
Finished with the introduction.

In this context, Huang et al.~\cite{DBLP:conf/apvis/Huang07,DBLP:journals/vlc/HuangEH14} a decade ago introduced some important statistical evidence (through a series of eye-tracking experiments), that edge crossings may not negatively affect the drawing's quality too much (and therefore the human's ability to read and interpret it), when the angles formed by the crossing edges are large. In other words, while prior to these experiments it was commonly accepted that mainly the number of crossings is the most important parameter for judging the quality and readability of a non-planar graph drawing, it turned out that the types of edge crossings also matter. As a result, a new and prominent research direction was initiated, which is unofficially recognized under the term ``beyond planarity''~\cite{Shonan2016,Dagstuhl2016,SoCG2017} and mainly focuses on graphs and their properties, when different constraints on the types of edges crossings are imposed; refer to~\cite{DBLP:journals/jgaa/BekosKM18,DBLP:journals/corr/abs-1804-07257} for recent surveys.

@@

From a practical point of view, we are only aware of two methods that aim in drawings with high crossing resolution; both of them are adjustments of spring embedding algorithms. The first one is due to Huang, Eades, Hong and Lin~\cite{DBLP:journals/vlc/HuangEHL13}, while the second one due to Argyriou, Bekos and Symvonis~\cite{DBLP:journals/cj/ArgyriouBS13}. Common in both algorithms are appropriate forces, which are applied on the endvertices of every pair of crossing edges. Of course, each of the two algorithms uses a different way to compute (the direction and the magnitude of) the forces, but the underlying idea in both algorithms is the same: the smaller the corresponding crossing angles are, the larger are the magnitudes of the forces applied at their endvertices.

In this paper, we attack the crossing resolution maximization problem by following a completely different approach. We suggest a simple and quite intuitive randomization method for computing drawings with high crossing resolution, which in practise turned out to be quite efficient\footnote{Note that, in general, randomization is a technique that has not been deeply examined in Graph Drawing, as it seems to be difficult to computed the expected quality of the produced drawings.}. However, since we could not provide any theoretical guarantee on the expected quality of the produced drawings (mainly due to the nature of the problem itself), we decided to follow a more practical approach. To this end, we implemented our algorithm and the ones presented in~\cite{DBLP:journals/vlc/HuangEHL13} and~\cite{DBLP:journals/cj/ArgyriouBS13}, and we experimentally compared the crossing resolutions of the drawings produced by them for standard benchmark graphs (such as the Rome and the North graphs) that are widely used in Graph Drawing for comparing different drawing algorithms. Our experimental evaluation indicates that our new method significantly outperforms the ones that are based on adjustments of spring embedding algorithms~\cite{DBLP:journals/vlc/HuangEHL13,DBLP:journals/cj/ArgyriouBS13} in terms of crossing resolution, but this comes at the cost of slightly higher running time, especially for large and dense graphs.

The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related results. In Section~\ref{sec:preliminaries}, we introduce preliminary notions and notation used throughout the paper. Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou at al.~\cite{DBLP:journals/cj/ArgyriouBS13} in Section~\ref{sec:experiments}. We conclude in Section~\ref{sec:conclusions} with open problems and future direction.

@@

@@

>>>>

In this context, Huang et al.~\cite{DBLP:conf/apvis/Huang07,DBLP:journals/vlc/HuangEH14} a decade ago introduced some important statistical evidence (through a series of eye-tracking experiments), that edge crossings may not negatively affect the drawing's quality too much (and therefore the human's ability to read and interpret it), when the angles formed by the crossing edges are large. In other words, while prior to these experiments it was commonly accepted that mainly the number of crossings is the most important parameter for judging the quality and readability of a non-planar graph drawing, it turned out that the types of edge crossings also matter. As a result, a new and prominent research direction was initiated, which is unofficially recognized under the term ``beyond planarity''~\cite{Shonan2016,Dagstuhl2016,SoCG2017} and mainly focuses on graphs and their properties, when different constraints on the types of edges crossings are imposed; refer to~\cite{DBLP:journals/corr/abs-1804-07257} for a recent survey.

@@

From a practical point of view, we are only aware of two methods that aim in drawings with high crossing resolution; both of them are adjustments of spring embedding algorithms~\cite{DBLP:journals/congnum/Eades84}. The first one is due to Huang, Eades, Hong and Lin~\cite{DBLP:journals/vlc/HuangEHL13}, while the second one due to Argyriou, Bekos and Symvonis~\cite{DBLP:journals/cj/ArgyriouBS13}. Common in both algorithms are appropriate forces, which are applied on the endvertices of every pair of crossing edges. Each of the two algorithms uses a different way to compute (the direction and the magnitude of) the forces, but the underlying idea in both algorithms is the same: the smaller the corresponding crossing angles are, the larger are the magnitudes of the forces applied at their endvertices.

In this paper, we attack the crossing resolution maximization problem by following a completely different approach. We suggest a simple and quite intuitive randomization method for computing drawings with high crossing resolution, which in practise turned out to be quite efficient\footnote{Note that, in general, randomization is a technique that has not been deeply examined in Graph Drawing, as it seems to be difficult to computed the expected quality of the produced drawings. A notable exception is the randomized approach by Goldschmidt and Takvorian~\cite{DBLP:journals/networks/GoldschmidtT94} for computing large planar subgraphs.}. However, since we could not provide any theoretical guarantee on the expected quality of the produced drawings (mainly due to the nature of the problem itself), we decided to follow a more practical approach. To this end, we implemented our algorithm and the ones presented in~\cite{DBLP:journals/vlc/HuangEHL13} and~\cite{DBLP:journals/cj/ArgyriouBS13}, and we experimentally compared the crossing resolutions of the drawings produced by them for standard benchmark graphs (such as the Rome and the North graphs) that are widely used in Graph Drawing for comparing different drawing algorithms. Our experimental evaluation indicates that our new method significantly outperforms the ones that are based on adjustments of spring embedding algorithms~\cite{DBLP:journals/vlc/HuangEHL13,DBLP:journals/cj/ArgyriouBS13} in terms of crossing resolution, but this comes at the cost of slightly higher running time, especially for large and dense graphs.

The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related works. In Section~\ref{sec:preliminaries}, we introduce preliminary notions and notation used throughout the paper. Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou at al.~\cite{DBLP:journals/cj/ArgyriouBS13} in Section~\ref{sec:experiments}. We conclude in Section~\ref{sec:conclusions} with interesting open problems and future~directions.

@@

As already mentioned, the study of the crossing resolution maximization problem has mainly focused on (propertied of) RAC graphs, that is, on the optimal case of the crossing resolution maximization problem. The study was initiated by Didimo et al.~\cite{DBLP:journals/tcs/DidimoEL11}, who showed that an $n$-vertex RAC graph has at most $4n-10$ edges. 
Angelini et al.~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11} presented results for two interesting variants, in which the input graph is either an acyclic digraph or a graph of bounded vertex degree. Argyriou et al.~\cite{DBLP:journals/jgaa/ArgyriouBS12} showed that the problem of deciding whether a graph is RAC is NP-hard. Didimo et al.~\cite{DBLP:journals/ipl/DidimoEL10} completely characterized the complete bipartite RAC graphs. Di~Giacomo et al.~\cite{DBLP:journals/algorithmica/GiacomoDEL14}, and Hong and Nagamochi~\cite{DBLP:conf/wg/HongN15} studied variants, in which the vertices of the input graph are restricted on two parallel lines and on the circumference of a circle, respectively. Eades and Liotta~\cite{DBLP:journals/dam/EadesL13} showed that the optimal RAC graphs (i.e., those of maximum density) are 1-planar (i.e., they admit drawings in which each edge is crossed at most once). Deciding, however, whether a 1-planar graph is RAC is NP-hard~\cite{DBLP:journals/tcs/BekosDLMM17}. Other relationships between the class of RAC graphs and subclasses of 1-planar graphs have also been studied by Bachmaier et al.~\cite{DBLP:journals/dam/BachmaierBHNR17} and Brandenburg et al.~\cite{DBLP:journals/tcs/BrandenburgDEKL16}. Note that the problem of finding RAC drawings has been also studied in the presence of bends; see e.g.~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/comgeo/ArikushiFKMT12,DBLP:journals/tcs/DidimoEL11,DBLP:journals/mst/GiacomoDLM11}.

To the best of our knowledge, there is only one work, by Dujmovic et al.~\cite{DBLP:journals/cjtcs/DujmovicGMW11}, which studies the crossing resolution maximization problem by relaxing the right-angle constraint on the crossing angles of the computed drawings. More precisely, in their work, Dujmovic et al.~\cite{DBLP:journals/cjtcs/DujmovicGMW11} proved that an $n$-vertex graph, whose crossing resolution is at least $\alpha$ has at most $(3n-6)\pi/\alpha$ edges. Corresponding density results for the case, in which few bends are allowed along each edge, are known by Ackerman et al.~\cite{DBLP:journals/siamdm/AckermanFT12} and Di Giacomo et al.~\cite{DBLP:journals/mst/GiacomoDLM11}.

An immediate observation emerging from the above literature overview is that the focus of the study of the crossing resolution maximization problem has been primarily on theoretical aspects of the problem (as naturally expected, of course). The absence of approaches for computing drawings with high crossing resolution in practice is rather obvious and it can be justified, up to a certain point, by fact that the problem turns to be NP-hard, when asking for drawings with optimal crossing resolution. Notable exceptions are of courses the works of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou at al.~\cite{DBLP:journals/cj/ArgyriouBS13}, which are both based on the spring-embedding technique~\cite{DBLP:journals/congnum/Eades84}. According to this technique, a graph is modelled as a physical system with forces acting on it, and a (good) drawing is obtained by an equilibrium state of the system; for an introduction and a detailed discussion of several different variants refer to~\cite{DBLP:books/ph/BattistaETT99}.

@@

Unless otherwise specified, in this paper we consider simple undirected graphs. Let $G=(V,E)$ be such a graph. The degree of vertex $u\in V$ of $G$ is denoted by $d(u)$. The degree $d(G)$ of  graph $G$ is defined as the maximum degree of its vertices, i.e., $d(G)=\max_{u\in V}d(u)$.

Given a drawing $\Gamma(G)$ of $G$ on the plane, we denote by $p_u=(x_u,y_u)$ the position of vertex $u \in V$ of $G$ in $\Gamma(G)$. For a pair of vertices $u,v\in V$ of $G$, the unit-length vector from $p_u$ to $p_v$ is denoted as $\overrightarrow{p_up_v}$, while the line segment delimited by $p_u$ and $p_v$ as $\overline{p_up_v}$.
